---
title: "Multiple Linear and Nonlinear Regression"
subtitle: "Module 11"
author: "Caio dos Santos"
date: "2026-04-12"
categories: [statistics]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

# Introduction

In Module 10, we kept things simple by design: one predictor, one
response, and a straight line. That simplicity was intentional. It let
us focus on the core habits that actually matter: starting with a
biological question, checking whether the model shape makes sense, and
separating statistical significance from practical reality.

But real agronomy is seldom that simple.

Think about corn yield for a moment. How much does it depend on rainfall
alone? Maybe some. But yield also responds to many factors, like soil
texture, organic matter, pH, temperature, and management choices. All of
these act together.

And here's the thing: many of those relationships are not straight
lines. They curve. They plateau. They follow S-shaped patterns.

So Module 11 takes the habits from Module 10 and extends them to handle
two real-world challenges:

-   **Multiple regression**: When you have several predictors, not just
    one
-   **Nonlinear modeling**: When the relationship is curved, not
    straight

The core question doesn't change:

> What model structure gives interpretable, biologically reasonable, and
> practically useful predictions?

We're just using better tools to answer it.

# Part 1: Multiple Linear Regression

## The Model: From One Predictor to Many

**Here's the key idea**: We're going to keep the same basic model
structure, but expand it from one predictor to many.

In Module 10, we looked at this simple model:

$$Y = \beta_0 + \beta_1 X + \varepsilon$$

Now we add more predictors:

$$
Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_p X_p + \varepsilon
$$

The algebra looks the same. But here's where things get interesting: the
interpretation changes fundamentally.

**In simple regression**, $\beta_1$ is straightforward:

> "For every one-unit increase in $X$, $Y$ changes by $\beta_1$ units"

**In multiple regression**, each coefficient tells a different story:

> "For every one-unit increase in $X_j$, $Y$ changes by $\beta_j$ units
> *after accounting for all the other predictors.*"

Let me say that again because it's crucial: **after accounting for all
the other predictors.** This is what makes multiple regression powerful
and also what makes it tricky to interpret.

When you have multiple predictors, each coefficient is what we call a
**partial effect**. It's the association between one predictor and the
response after adjusting for the other predictors in the model.

This is a conditional association, not automatically a causal effect.

Here is a concrete illustration to make this real:

Suppose two counties in Minnesota receive exactly the same amount of
rainfall. But one has higher organic matter in its soil than the other.
If the model gives $\beta_{om} = 0.8$, here's what that means:

> Holding rainfall (and all other variables) constant, yield increases
> by 0.8 units for each one-unit increase in OM

Notice what we're doing here: we're comparing two counties that are
**similar in everything except organic matter**.

We're not comparing "all high-OM counties" vs "all low-OM counties"
(because those would differ in rainfall too, and a thousand other
things).

That's the whole point. This is what makes it a **partial** effect.

**This does NOT mean**: We have created a controlled experiment. We
didn't go out and manipulate organic matter while holding rainfall
steady. We're just using the model to statistically adjust for the
differences.

This adjusted interpretation is why multiple regression matters. Without
it, the key question can't be answered: "Is this factor actually
important, or is it just correlated with something else that matters?"

## Case Study: Which Soil and Climate Factors Actually Drive Corn Yield?

Here's a concrete scenario that makes this necessary. Imagine you're a
soil scientist working with a cooperative extension office. A farmer
walks in and asks:

> "I know rainfall matters for yield, but in this region, which soil
> properties matter most? Is it organic matter? Texture? pH?"

That's a great question. But here's why you can't just run three
separate regressions:

Across counties, rainfall, organic matter, and soil texture are all
correlated. Areas with high organic matter typically also have better
rainfall. If you regressed yield on OM alone, you'd see a big positive
effect. But is that really because of OM, or is it just because those
counties happen to have better rainfall too?

**This is where multiple regression helps.** Instead of asking "Does OM
matter?" ask the real question: "After accounting for rainfall, does OM
still explain variation in yield?"

Let's work through it with county-level data from Minnesota and
Wisconsin.

```{r}
library(ggplot2)

counties <- read.csv("data/county_environments.csv")
head(counties)
```

For this example, the analysis focuses on Minnesota and Wisconsin.

```{r}
counties_mn_wi <- subset(counties, state %in% c("MN", "WI"))

vars <- c("corn", "ppt", "sand", "clay", "om", "sph")
model_data <- na.omit(counties_mn_wi[, vars])

dim(model_data)
summary(model_data)
```

## Plot the Data First

Before fitting anything, inspect the data. No exceptions.

```{r}
plot(model_data, pch = 20, cex = 0.7)
```

A scatterplot matrix isn't pretty, but it's honest. You'll see which
variables move together, which relationships look straight or curved,
and how much your predictors are correlating with each other. Your eyes
will catch patterns that a test will miss.

## Start with a Simple Model (Two Predictors)

Let's start with a simple model using only rainfall and organic matter
to predict corn yield. This might feel like a stripped-down model, but
that's exactly the point. Simplicity helps us see the core logic before
we add complexity.

```{r, echo=TRUE}
model_simple <- lm(corn ~ ppt + om, data = model_data)
summary(model_simple)
```

So you've got the output. Now what? Here's what NOT to do: don't just
scan for low p-values and move on.

Instead, scrutinize each coefficient carefully. First, the **sign**:
does it match biology? Does more rainfall increase yield? That makes
sense. If rainfall decreases yield, that's a red flag unless we're
talking about poorly drained soils.

Then check **uncertainty**. Is the p-value small, and do the confidence
intervals exclude zero? Wide confidence intervals mean the model doesn't
really know where the effect lives. Finally, **plausibility**: Given
your data and model assumptions, does this make biological sense?
Sometimes you get a technically valid result that's biologically
nonsensical.

### Visualizing Partial Effects

If graphs help you, here's a direct way to see partial effects.

Let's use a stripped-down teaching model with just two predictors:
rainfall and organic matter.

$$
\text{corn} = \beta_0 + \beta_{ppt}\,\text{ppt} + \beta_{om}\,\text{om} + \varepsilon
$$

This makes the partial-effect idea easier to see without extra moving
parts. In this setup, $\beta_{om}$ is the change in predicted yield for
a one-unit increase in OM **after accounting for rainfall**.

```{r, include=FALSE}
om_levels <- quantile(model_data$om, probs = c(0.1, 0.9), na.rm = TRUE)

pred_grid <- expand.grid(
  ppt = seq(min(model_data$ppt), max(model_data$ppt), length.out = 200),
  om = as.numeric(om_levels)
)

pred_grid$om_group <- factor(
  pred_grid$om,
  levels = as.numeric(om_levels),
  labels = c("Lower OM (10th percentile)", "Higher OM (90th percentile)")
)

pred_grid$corn_pred <- predict(model_simple, newdata = pred_grid)

# anchor the annotation at the midpoint of the ppt range
arrow_x    <- median(model_data$ppt, na.rm = TRUE)
arrow_ppt  <- data.frame(ppt = arrow_x, om = as.numeric(om_levels))
arrow_y    <- predict(model_simple, newdata = arrow_ppt)
arrow_ylo  <- arrow_y[1]
arrow_yhi  <- arrow_y[2]
arrow_ymid <- mean(c(arrow_ylo, arrow_yhi))

partial_effect_plot <- ggplot() +
  geom_point(
    data = model_data,
    aes(x = ppt, y = corn),
    color = "gray65",
    alpha = 0.35,
    size = 1.7
  ) +
  geom_line(
    data = pred_grid,
    aes(x = ppt, y = corn_pred, color = om_group),
    linewidth = 1.25
  ) +
  # double-headed arrow spanning the gap between lines
  annotate(
    "segment",
    x = arrow_x, xend = arrow_x,
    y = arrow_ylo, yend = arrow_yhi,
    arrow = arrow(ends = "both", length = unit(0.18, "cm"), type = "closed"),
    color = "gray30", linewidth = 0.7
  ) +
  # label in the middle of the arrow
  annotate(
    "text",
    x = arrow_x + 19, y = arrow_ymid,
    label = 'OM effect',
    size = 3.8, label.size = 0, fill = "white", color = "gray20"
  ) +
  scale_color_manual(values = c("#1b9e77", "#d95f02")) +
  labs(
    title = "Partial Effects with a Two-Predictor Model",
    subtitle = "Predicted lines from corn ~ ppt + om",
    x = "Rainfall (ppt)",
    y = "Predicted corn yield",
    color = "Organic matter"
  ) +
  theme_minimal(base_size = 12)
```

```{r, echo=FALSE}
partial_effect_plot
```

How to read it quickly:

-   At any fixed rainfall, the vertical gap between the two lines is the
    OM partial effect.
-   The lines are parallel because there is no interaction term
    (`ppt:om`). That means the OM effect is constant across rainfall
    values in this model.

**Pause and reflect:** If two counties have the same rainfall but
different OM, the model-predicted yield gap between those counties is
the partial OM effect.

### Expand to the Full Model

Once that logic is clear, we can expand to the full model with
additional predictors:

```{r, echo=TRUE}
model_full <- lm(corn ~ ppt + sand + clay + om + sph, data = model_data)
summary(model_full)
```

This is the same idea as before, just with more variables being adjusted
for at the same time. You're not changing the logic, only increasing
realism.

## Check Assumptions

Just like in Module 10, we need to look at the diagnostics. Here's what
these four plots are actually telling you:

```{r, echo=TRUE}
par(mfrow = c(1, 2))
plot(model_full, which = 1:2)
par(mfrow = c(1, 1))
```

**Residuals vs Fitted**: You want to see a random cloud centered around
zero. No patterns. If it fans out (wider at high values than low),
that's heteroscedasticity—your model loses confidence at some
predictions. If you see a U-shape or wave, the model is systematically
biased.

**Normal Q-Q**: Points should track the diagonal line. Small wiggles at
the tails are usually fine in large samples, but a systematic S-shape
means residuals aren't normally distributed. That's worth noting. In
larger samples, this is usually more about inference precision (p-values
and intervals) than mean-prediction quality.

**Quick check:** If at least one diagnostic shows a clear pattern, pause
before interpreting coefficients as if everything is fine.

### What's Going On When Coefficients Change?

Suppose you fit a model with both sand and clay. You get a negative
coefficient on sand. Then you remove clay from the model and refit. Now
sand has a positive coefficient. Something flipped.

Your first thought might be: "That's weird. Did I make a mistake?"

**Not necessarily.** This is actually a prediction.

Here's what's happening:

Sand and clay are correlated in most cases (areas with more sand have
less clay). When clay is in the model, the model is extracting the
partial effect of sand **after already accounting for clay**. When you
remove clay, sand's job changes. Now sand has to explain more variation
because clay isn't there to help. So the sign can flip.

**What's the lesson here?** This is a consequence of
**multicollinearity**—when predictors are correlated, they can share or
reallocate explained variation among themselves. It's one reason why you
have to be careful when interpreting individual coefficients.

## Diagnose Multicollinearity

Here's the problem: If two predictors are highly correlated, their
partial effects become **unstable**. A small change in the data can flip
the sign or magnitude of one of the coefficients. This is bad because
you can't trust what the coefficient is telling you.

The correlation matrix is a good first check:

```{r, echo=TRUE}
predictor_cor <- cor(model_data[, c("ppt", "sand", "clay", "om", "sph")])
round(predictor_cor, 2)
```

Look for pairs of predictors with correlations near 0.7 or higher (in
absolute value). **This is not an absolute cutoff** but it's a good
indication that standard errors could increase, which typically leads to
larger p-values and wider confidence intervals. That makes real effects
harder to detect as predictors compete for the same variation.

**Pause and reflect:** if your goal is prediction, correlated predictors
can still help. If your goal is interpretation, correlation becomes a
much bigger problem.

Does that automatically mean you need to remove a predictor? Not
necessarily. But it absolutely means you need to interpret carefully.
Not all multicollinearity is fatal, but you have to know when you're
dealing with it.

### Worked Example: Variance Inflation Factor (VIF)

The correlation matrix is useful, but here's an even more direct
measure: the **Variance Inflation Factor**. It tells you how much the
uncertainty of a coefficient is inflated due to correlations with other
predictors.

Here's the idea: Imagine you regressed one predictor (say, sand) on all
the other predictors. If you can predict sand really well from the other
variables, that means sand is redundant. This is because it's explaining
variation that's already covered. In that case, sand's coefficient will
be uncertain, and its standard error will increase, which usually pushes
p-values upward.

The math is:

$$
VIF_j = \frac{1}{1 - R_j^2}
$$

where $R_j^2$ comes from regressing $X_j$ on the other predictors.

Here's how to interpret VIF: values around 1 mean minimal
collinearity—the predictor is doing its own thing. Values above about 5
suggest meaningful collinearity and loss of precision. Values above
about 10 are often considered problematic. These are heuristics, not
hard rules, so context still matters.

**Watch out for compositional variables.** Soil texture is the classic
trap: sand, silt, and clay must sum to 100% by definition. That's not
collinearity—that's *perfect* linear dependence. Your VIF will be
infinite. This is a special case that requires special handling.

Here's how it plays out with a worked example:

```{r, include=FALSE}
vif_manual <- function(df) {
  out <- sapply(names(df), function(var) {
    others <- setdiff(names(df), var)
    form <- as.formula(paste(var, "~", paste(others, collapse = " + ")))
    r2 <- summary(lm(form, data = df))$r.squared
    if (isTRUE(all.equal(r2, 1, tolerance = 1e-10))) {
      Inf
    } else {
      1 / (1 - r2)
    }
  })
  round(out, 2)
}
```

Let's run VIF on two scenarios:

Let's include all three soil texture classes as predictors in a model
along with organic matter, and let's take a look at the summary of this
model.

```{r, echo=TRUE}
model_inflated <- lm(corn ~ sand + silt + clay + om, data = counties_mn_wi)
summary(model_inflated)
```

Estimates and standard errors for all three predictors seem to be
inflated, right? Let's take a look at the VIF for these predictors. We
will use the `vif()` function in the library called `car`, which stands
for *Companion to Applied Regression*.

```{r, echo=TRUE}
library(car)
vif(model_inflated)
```

We were talking about rules of thumb and saying that values above 10
were problematic... Those values are pretty extreme.

Let's get rid of one of the soil texture classes, look at the summary
and compute the VIF again.

```{r, echo=TRUE}
model_adjusted <- lm(corn ~ sand + clay + om, data = counties_mn_wi)
summary(model_adjusted)
```

The estimates for the effects of sand and clay seem a little more
plausible, also, the standard error of the effect of these two
predictors seems way less inflated. Let's compute the VIF to check.

```{r, echo=TRUE}
vif(model_adjusted)
```

The VIF looks more under control now, indicating that keeping sand and
clay as predictors could work.

**What's happening here?**

In the first case, we included all three texture fractions—sand, silt,
and clay. The VIF values are huge for the texture variables. Why?
Because they're compositional: they sum to 100% by definition. This
means if you know sand and clay, you automatically know silt (it's 100%
minus the others). So the model can't distinguish between their effects.
This is **perfect linear dependence**.

In the second case, we dropped one texture variable (silt). Now the VIF
values are finite and interpretable: sand ≈ 3.5, clay ≈ 3.3, OM ≈ 1.1.

With compositional data, you have to drop one category to avoid perfect
collinearity. Once you do that, you can estimate the partial effects
properly.

## Keep It as Simple as Possible (But Not Simpler)

Here's a temptation: the more predictors you add to a model, the higher
$R^2$ gets. It always does. So why not just throw everything in?

Because **more is not always better.**

A simpler model is easier to interpret, easier to communicate to
stakeholders, and less prone to overfitting. So the real question is: Do
the extra predictors earn their place?

This is a decision step, not a math contest.

Let's compare a full model (all five predictors) with a reduced model
(dropping clay):

```{r, echo=TRUE}
model_reduced <- lm(corn ~ ppt + sand + om + sph, data = model_data)

summary(model_reduced)
anova(model_reduced, model_full)
AIC(model_reduced, model_full)
```

Now look at the results together. Here's what each number is telling
you:

**ANOVA p-value**: If this is large (\> 0.05), it says "clay doesn't add
much once you've already got the other variables." The reduced model
explains essentially the same variation.

(This comparison assumes the models are nested and fit on the same
dataset with identical observations.)

**AIC (Akaike Information Criterion)**: Lower is better. But AIC
penalizes complexity, so a simpler model can have lower AIC even if
$R^2$ is slightly smaller. Think of it as "How much better would I have
to get if I'm going to justify adding more complexity?"

**Adjusted** $R^2$: This penalizes unnecessary predictors. Compare this
between models, not the raw $R^2$.

So how do you decide in practice?

-   ANOVA p-value is large (the dropped predictor adds little)
-   AIC is similar or lower for the reduced model

**Lesson:** model selection is a biological and statistical decision,
not a race for the highest $R^2$.

# Part 2: Nonlinear Relationships

Straight lines are convenient for teaching but biology rarely
cooperates.

So here's the transition from Part 1: once a linear model starts missing
structure, we need a model that matches biology better.

Think about nitrogen response in corn. At low N rates, each additional
kg gives a big yield bump. But keep adding N and the returns shrink.
Eventually more N does nothing. That's not a straight line—it's curved.

Or think about plant growth over time. Early on, growth is slow. Then it
accelerates. Then it slows again as the plant matures. That's an
S-curve, not a line.

When you fit a straight line to curved data, your residuals will show
systematic patterns. That's your signal: **It's time to move beyond
linear.**

Here's the diagnostic approach—and notice it's the same habit from
Module 10:

**Look at the residuals.**

Nonlinear modeling isn't a different workflow. It's what you do *after*
the linear model shows clear problems in the data.

When does a linear model genuinely fail? Here are the warning signs:

1.  **Residual plots show systematic curvature**. Not random scatter,
    but a U-shape or wave pattern. The model over-predicts at low values
    and under-predicts at high values (or vice versa). That's your
    signal that the true relationship is curved.

2.  **Predictions become biologically nonsensical**. A linear model for
    plant growth might predict negative biomass at early time points. A
    nitrogen response model might predict infinite yield at very high N
    rates. That's not realistic, and it signals that curvature matters.

3.  **You know the mechanism is curved**. Sometimes domain knowledge
    tells you the relationship must follow an S-curve (logistic growth)
    or a linear-plateau (nutrient saturation). Don't ignore that
    knowledge just because a line "fits."

Now comes the choice:

There are two main strategies, and they're useful in different
situations:

-   **Strategy 1**: Stay in a linear framework but add polynomial terms
    (quadratic, cubic, etc.). This keeps fitting simple.
-   **Strategy 2**: Fit a nonlinear model directly with a biological
    curve shape built in. This requires more care and can extrapolate
    more sensibly when the chosen form matches the underlying mechanism.

**Pause and reflect:** if you care most about a quick local fit, start
with polynomial terms. If you care most about biologically meaningful
parameters and realistic extrapolation, go nonlinear.

## Strategy 1: Polynomial Regression in a Linear Framework

### When to use polynomials

Polynomials (squared terms, cubic terms, etc.) are attractive because:

-   They stay in the linear regression framework, so all the tools you
    know work
-   They're easy to fit
-   You can use all the diagnostics from Module 10

They work well when you need to capture curvature but don't have strong
prior knowledge about what that curve should look like.

### Example: Nitrogen Response with Curvature

Look at a simple linear fit to nitrogen response:

```{r, echo=TRUE}
corn_n <- read.csv("data/corn_nrate_mono.csv")

ggplot(corn_n, aes(x = n_rate, y = yield)) +
  geom_point(pch = 21, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "tomato") +
  labs(x = "N rate", y = "Yield") +
  theme_bw()
```

The straight line captures the general trend, but the response is
clearly curved. It's steep at low N and flattens out at high N. Adding a
squared term captures this bend:

```{r, echo=TRUE}
model_poly <- lm(yield ~ n_rate + I(n_rate^2), data = corn_n)
summary(model_poly)
```

The model is:

$$
Y = \beta_0 + \beta_1 X + \beta_2 X^2 + \varepsilon
$$

When $\beta_2$ is negative (like here), the curve bends downward and
represents **diminishing returns**—each additional unit of N produces
less gain.

Here's the crucial part: **Don't interpret** $\beta_1$ alone. When a
squared term is present, the effect of $X$ depends on both $\beta_1$ and
$\beta_2$. The marginal effect of $X$ is $\beta_1
+ 2\beta_2 X$, so it depends on the value of $X$ (i.e., the slope
changes along the curve).

**Quick check:** at low N and high N, does the model imply the same
slope? If not, that's exactly why curvature matters.

The best way to understand the effect? Make a prediction plot:

```{r, echo=FALSE}
grid_poly <- data.frame(n_rate = seq(min(corn_n$n_rate), max(corn_n$n_rate), length.out = 200))
grid_poly$pred <- predict(model_poly, newdata = grid_poly)

ggplot(corn_n, aes(x = n_rate, y = yield)) +
  geom_point(pch = 21, size = 2) +
  geom_line(data = grid_poly, aes(y = pred), color = "steelblue", linewidth = 1) +
  labs(x = "N rate", y = "Yield") +
  theme_bw()
```

The curve shows what's actually happening: steep initial response, then
flattening. That's what matters. That's more meaningful than any
coefficient you could report.

## Strategy 2: Fit Nonlinear Directly with nls()

### When to use nonlinear models

Nonlinear models are powerful when:

-   You understand the mechanism and can describe the curve shape
    (monomolecular, logistic, etc.)
-   Parameter values have biological meaning and you want to report them
-   You need more control over extrapolation (a line predicts infinite
    growth; a plateau model doesn't)

The tradeoff is complexity. Nonlinear models require careful fitting,
diagnostics, and validation.

### Example: Nutrient Response with a Plateau

At low N rates, yield jumps with each additional kilogram. But
eventually, adding more N doesn't help—the yield plateaus.

A straight line would predict infinite yield at extreme N rates. That's
nonsense. What we need is a model that rises linearly at first, then
levels off. That's exactly what the **linear-plateau** model does.

For this, we will use the `nlraa` library, which stands for *Nonlinear
Regression for Agricultural Applications*. If you want to know more
about this, I highly recommend the manuscript [Nonlinear Regression
Models and Applications in Agricultural
Research](https://doi.org/10.2134/agronj2012.0506), written by Sotirios
Archontoulis and Fernando Miguez.

```{r, echo=TRUE}
library(nlraa)
corn_n <- read.csv("data/corn_nrate_mono.csv")

ggplot(corn_n, aes(x = n_rate, y = yield)) +
  geom_point(pch = 21, size = 2) +
  labs(x = "N rate", y = "Yield") +
  theme_bw()
```

Now fit the linear-plateau model:

```{r, echo=TRUE}
model_linp <- nls(yield ~ SSlinp(n_rate, a, b, xs), data = corn_n)
summary(model_linp)
```

What do these parameters mean? **This is the kind of thing that makes
nonlinear models powerful**:

-   **a**: The intercept—predicted yield at zero N.
-   **b**: The slope in the linear phase—each additional kg of N adds
    this much yield, up to the join point.
-   **xs**: The join point—the N rate at which the yield stops
    responding and plateaus. This is often close to the agronomic
    optimum and can inform economic decisions, but the true economic
    optimum depends on crop price, fertilizer cost, and risk.

The fitted curve:

```{r}
grid_n <- data.frame(n_rate = seq(min(corn_n$n_rate), max(corn_n$n_rate), length.out = 200))
grid_n$pred <- predict(model_linp, newdata = grid_n)

ggplot(corn_n, aes(x = n_rate, y = yield)) +
  geom_point(pch = 21, size = 2) +
  geom_line(data = grid_n, aes(y = pred), color = "steelblue", linewidth = 1) +
  labs(x = "N rate", y = "Yield") +
  theme_bw()
```

See how the curve rises steeply then flattens sharply at the join point?
The `xs` parameter tells you exactly where that happens. That's the kind
of number you can hand to a farmer.

**Lesson:** polynomial models can describe shape, but mechanistic
nonlinear models often give you parameters you can act on.

### Example: Plant Growth Following an S-Curve

Now a different scenario: a crop physiologist tracking soybean biomass
from emergence to maturity.

Early on, growth is slow (small plant, limited leaf area). Then it
accelerates—the plant has more leaves photosynthesizing, growth
compounds. Finally, it slows again as the plant matures and stops
investing in biomass accumulation.

That pattern is an **S-curve** (sigmoidal). It's not a line, and a
polynomial might fit it, but an S-curve model will tell you when growth
is fastest (the inflection point), which is biologically interesting.

```{r}
soy_tdm <- read.csv("data/soybean_dae_tdm_logis.csv")

ggplot(soy_tdm, aes(x = dae, y = tdm)) +
  geom_point(pch = 21, size = 2) +
  labs(x = "Days after emergence", y = "Total dry matter") +
  theme_bw()
```

Fit a logistic (S-curve) model:

```{r, echo=TRUE}
model_logis <- nls(tdm ~ SSlogis(dae, Asym, xmid, scal), data = soy_tdm)
summary(model_logis)
```

Again, notice the parameters have meaning:

-   **Asym**: Maximum biomass the plant ever reaches
-   **xmid**: The day when growth is fastest (inflection point of the
    S-curve)
-   **scal**: How steeply the curve rises. Bigger value = sharper
    transition

For a crop physiologist, xmid tells you when the plant is at peak growth
rate. Asym tells you about ultimate biomass potential. These are
actionable biological insights.

```{r}
grid_dae <- data.frame(dae = seq(min(soy_tdm$dae), max(soy_tdm$dae), length.out = 200))
grid_dae$pred <- predict(model_logis, newdata = grid_dae)

ggplot(soy_tdm, aes(x = dae, y = tdm)) +
  geom_point(pch = 21, size = 2) +
  geom_line(data = grid_dae, aes(y = pred), color = "darkgreen", linewidth = 1) +
  labs(x = "Days after emergence", y = "Total dry matter") +
  theme_bw()
```

## Polynomial vs Nonlinear: Which One Should You Use?

Use this decision rule:

-   If you need a quick flexible fit and a familiar workflow, start with
    polynomial terms.
-   If you need interpretable biological parameters (like join point or
    asymptote) and realistic extrapolation, use nonlinear models.

Neither is always best. The right choice depends on your question, your
data range, and whether you need explanation, prediction, or both.

# Common Pitfalls and How to Avoid Them

Let me call out the errors that tend to trap people:

**1. Confusing correlation with causation.**

Multiple regression finds patterns. It doesn't prove causation. A model
predicting yield from soil properties might just be picking up
confounding relationships. Domain knowledge and careful reasoning still
matter. Statistics is a tool, not a truth machine.

**2. Trusting** $R^2$ too much.

A model can explain 95% of variation in your training data and still
fail badly on new observations. That's overfitting. Always evaluate
likely performance on future or held-out data.

**3. Misinterpreting coefficients when multicollinearity is present.**

Correlated predictors make individual coefficients unstable. Check the
correlation matrix and VIF. If you see red flags, focus on what the
model actually predicts, not what individual coefficients say.

**4. Ignoring curved residuals.**

If your residual plot shows systematic curvature instead of random
scatter, the model is biased. Fix it with polynomial terms or nonlinear
models.

**5. Fitting a fancy nonlinear model without validation.**

Just because you can fit an S-curve doesn't mean you should. Check
whether the algorithm converged, whether parameter values make sense,
and whether different starting values give the same answer. If something
looks off, it probably is.

# Summary: What Changes and What Doesn't

Module 11 builds on Module 10. You're using more complex models, sure.
But the core thinking stays the same. And I want to emphasize this
because it's the real lesson: **The tools change. The reasoning
doesn't.**

Here's what every analysis should look like:

**1. Start with a biological question, not a dataset.**

There's a difference between "What variables predict yield?" and "I have
this dataset, let me analyze it." The first is research. The second is
busywork.

**2. Plot the data before fitting anything.**

No exceptions. Scatterplot matrices, histograms, everything. Your eyes
catch patterns that tests miss.

**3. Check assumptions.**

Residual plots, normality checks, heteroscedasticity. Are they perfect?
No. But do they tell a coherent story about what's working and what's
not? They should.

**4. Compare alternatives using statistics AND biology.**

AIC is useful. But also ask: "Does this make biological sense?" Both
matter equally.

**5. Pick the simplest model that's honest.**

Not the fanciest. Not the highest $R^2$. The simplest one that does real
work.

The core thinking is simple: **Think like a biologist first. Use
statistics to check whether your reasoning holds up, not as a substitute
for thinking.**

That's the part that actually matters.
