---
title: "Correlation and Simple Regression"
subtitle: "Module 10"
author: "Caio dos Santos"
date: "2026-04-03"
categories: [statistics]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

# Introduction

In the previous modules, we focused on comparing means across
qualitative treatments.

-   In modules 4 and 5, we compared two populations.
-   In modules 6 through 8, we compared multiple qualitative treatments.
-   In module 9, we discussed what to do when assumptions are not well
    behaved.

The same assumptions apply in regression. The difference is not in the
assumptions, but in how we model the mean: instead of a group mean, we
use a continuous function of $x$. Let's keep that continuity in mind
throughout this module.

In this module, we move to a different kind of question.

Instead of comparing only treatment groups identified by names (hybrid
A, fungicide B, treatment C), we now study relationships between
**quantitative variables**.

This is the big shift for units 10 and 11. In previous modules,
treatments were categorical: hybrids, fungicides, or herbicides with
names and labels. Those are different categories, but they are not
naturally arranged on a numeric scale. Here, we move to variables like
nitrogen rate, seed population, and canopy temperature. Because those
are numerical and ordered, we can ask a different question: not just
"are these groups different," but "how much does $y$ change as $x$
changes," and what do we predict between observed values?

Examples include:

-   nitrogen rate and yield
-   plant biomass and fruit count
-   canopy temperature and grain moisture

This module introduces two related but different tools:

-   **correlation**: measures association
-   **simple linear regression**: models how one variable changes as
    another changes

# Correlation versus Regression

When we analyze two quantitative variables, the first practical question
is:

> Are we only describing association, or are we modeling a directional
> relationship?

The clearest way to see the difference is visually. Both plots below use
the same example dataset. In one, we only summarize association. In the
other, we fit a directional model.

```{r}
library(ggplot2)

set.seed(99)
n_demo  <- 40
x_cd    <- rnorm(n_demo, 50, 12)
y_cd    <- 0.5 * x_cd + rnorm(n_demo, 0, 8)
demo_df <- data.frame(x = x_cd, y = y_cd)
r_cd    <- round(cor(demo_df$x, demo_df$y), 2)

ggplot(demo_df, aes(x, y)) +
  geom_point(pch = 21, size = 2) +
  annotate("text", x = min(demo_df$x) + 1, y = max(demo_df$y) - 1,
           label = paste0("r = ", r_cd),
           hjust = 0, size = 5, fontface = "bold") +
  labs(x = "x", y = "y",
       title = "Correlation: how strong is the association?") +
  theme_linedraw()
```

```{r}
ggplot(demo_df, aes(x, y)) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue") +
  geom_point(pch = 21, size = 2) +
  labs(x = "x", y = "y",
       title = "Regression: how does y change as x increases?") +
  theme_linedraw()
```

## Correlation

Correlation measures the strength and direction of association between
variables, usually represented as $r$.

-   $r$ ranges from -1 to 1
-   values near 1 indicate a strong positive association
-   values near -1 indicate a strong negative association
-   values near 0 indicate weak or no linear association

A key point:

> Correlation does not imply causation.

One important limitation: $r$ only measures **linear** association. A
strong curved relationship can still produce a low correlation
coefficient because the two halves of the curve partially cancel each
other out. Similarly, a single outlier can inflate $r$ substantially.
This means a high $r$ alone is not enough. Always plot the data first.

The example below illustrates this. Despite a strong, clear relationship
between $x$ and $y$, the correlation coefficient is near zero because
the curve is symmetric and the linear component cancels out.

```{r}
set.seed(7)
x_nl  <- seq(-3, 3, length.out = 50)
y_nl  <- -(x_nl ^ 2) + rnorm(50, 0, 0.4)
nl_df <- data.frame(x = x_nl, y = y_nl)
r_nl  <- round(cor(nl_df$x, nl_df$y), 2)

ggplot(nl_df, aes(x, y)) +
  geom_point(pch = 21, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue") +
  annotate("text", x = -1, y = -6,
           label = paste0("r = ", r_nl),
           hjust = 0, size = 4.5, fontface = "bold") +
  labs(x = "x", y = "y",
       title = "Strong nonlinear relationship, but r near 0") +
  theme_linedraw()
```

## Regression

Simple linear regression is used when we want to model how a response
variable $y$ changes with an explanatory variable $x$.

In plain terms, regression lets us do three practical things:

-   test whether $x$ is related to $y$
-   estimate the size of that relationship
-   predict $y$ for values of $x$ within the observed range

## Case Study 1: Cucumber Traits

An agronomist conducted a greenhouse experiment to understand the
relationship between plant architecture and early productivity in
cucumber. She grew 60 plants from the same seed lot under identical
conditions and, at the flowering stage, harvested and dissected each
plant to count:

-   number of main stem branches
-   total number of leaves
-   number of fruits that had set by the flowering stage ("early
    fruits")

Her goal was to identify which traits correlate with early fruiting,
which might help her breed for faster-producing varieties.

Here are the first few rows of the resulting dataset:

```{r, echo=FALSE}
library(ggplot2)
library(dplyr)

set.seed(20260403)
n <- 60

branches <- round(rnorm(n, mean = 65, sd = 18))
branches[branches < 10] <- 10

leaves <- round(8 * branches + rnorm(n, mean = 0, sd = 90))
leaves[leaves < 20] <- 20

earlyfruit <- round(0.22 * branches + rnorm(n, mean = 0, sd = 4))
earlyfruit[earlyfruit < 0] <- 0

cucumber <- data.frame(branches, leaves, earlyfruit)
head(cucumber)
```

Let us first visualize branches versus leaves.

```{r}
ggplot(cucumber, aes(x = branches, y = leaves)) +
  geom_point(pch = 21, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue") +
  labs(x = "Branches per plant",
       y = "Leaves per plant") +
  theme_linedraw()
```

Now branches versus early fruit.

```{r}
ggplot(cucumber, aes(x = branches, y = earlyfruit)) +
  geom_point(pch = 21, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick") +
  labs(x = "Branches per plant",
       y = "Early fruit count") +
  theme_linedraw()
```

In both plots, we see positive association. But this alone does not tell
us whether one variable causes the other.

For example, do more branches enable more leaves to form, or does more
active photosynthesis from abundant leaves fuel branch growth?
Similarly, in indeterminate crops like tomatoes or soybeans, branches
and fruits compete for the same carbohydrates, making the causal
direction ambiguous. It's also possible that a third variable, like
total plant vigor, drives both branches and fruits upward independently.
Correlation alone cannot resolve these questions.

## Measuring correlation with $r$

```{r, echo=TRUE}
cor(cucumber$branches, cucumber$leaves)
cor(cucumber$branches, cucumber$earlyfruit)
```

That gives us a quick numerical summary of linear association strength.

## What does $r$ measure?

One way to think about $r$ is that it takes covariance and puts it on a
common scale.

$$
r = \frac{\mathrm{cov}(x, y)}{s_x s_y}
$$

-   $\mathrm{cov}(x, y)$ captures how $x$ and $y$ vary together
-   $s_x$ and $s_y$ scale by the variability of each variable

That scaling step is why $r$ always stays between -1 and 1.

You do not need to memorize the formula. What matters most is:

-   sign of $r$ gives direction of association
-   magnitude of $r$ gives strength of linear association

A geometric way to picture this: imagine centering both $x$ and $y$
around their means so the origin sits at $(\bar{x}, \bar{y})$. If points
are aligned at approximately a 45-degree angle (northeast to southwest),
their products accumulate with the same sign, driving $r$ toward
$\pm 1$. If points scatter randomly across all quadrants, positive and
negative products cancel out, pushing $r$ toward 0. This is why the
denominator (which measures the maximum possible alignment) keeps $r$
bounded between -1 and 1: it is scaling by the theoretical maximum
covariance.

A natural bridge to regression is:

$$
\hat{\beta}_1 = r \cdot \frac{s_y}{s_x}
$$

This tells us that slope depends on both the correlation between $x$ and
$y$ **and** the scales of the two variables. Two datasets with the same
$r$ but different standard deviations will produce different slopes.
Keeping this in mind helps avoid the common mistake of equating
correlation magnitude with regression effect size.

## A short "by hand" correlation example in R

Up to this point, we have mostly used `cor(x, y)` and taken the result
at face value. That works perfectly fine, but it can feel a little like
a black box.

So let’s slow down for a minute and actually see what is going on under
the hood.

Suppose we have four paired observations:

```{r, echo=TRUE}
x <- c(2, 4, 5, 7)
y <- c(3, 6, 7, 10)

xbar <- mean(x)
ybar <- mean(y)

xbar
ybar
```

First step: center both variables by subtracting their means. This just
moves the data so that everything is measured relative to the average.

```{r, echo=TRUE}
x_centered <- x - xbar
y_centered <- y - ybar

data.frame(x, y, x_centered, y_centered)
```

Now here is the key idea.

-   When $x$ is above average and $y$ is also above average → positive
    contribution
-   When $x$ is below average and $y$ is also below average → also
    positive
-   When one is above and the other is below → negative contribution

So what we do is multiply these centered values together and add them
up.

```{r, echo=TRUE}
numerator <- sum(x_centered * y_centered)
numerator
```

That gives us a measure of how the variables move together. But we are
not done yet, because this value depends on the scale of the data.

To fix that, we divide by the variability in each variable.

```{r, echo=TRUE}
denominator <- sqrt(sum(x_centered ^ 2) * sum(y_centered ^ 2))
denominator
```

Now we put it all together:

```{r, echo=TRUE}
r_manual <- numerator / denominator
r_manual

cor(x, y)
```

So this is all correlation is doing.

It is checking whether observations that are above average in $x$ tend
to line up with observations that are above average in $y$, and the same
for below-average values.

Once you see it this way, $r$ stops being “just a number R gives you”
and becomes a summary of how consistently the two variables move
together.

# Regression

Now we move from association to modeling a directional relationship.

## Case Study 2: Nitrogen Rate and Corn Yield

A crop consultant evaluated corn's response to nitrogen fertilization on
a farmer's field. The farmer was considering three fertilizer programs:
no applied nitrogen (baseline), a moderate rate (about 34 lbs/acre), and
a high rate (about 67 lbs/acre), each applied to four randomly selected
plots. At harvest, the consultant measured grain yield (standardized to
15% moisture) from each plot.

The goal was to estimate the relationship between nitrogen rate and
yield so the farmer could make an economically informed decision about
the optimal application rate.

Here are the first few rows of the harvest results:

```{r}
set.seed(20260403)

nitro <- rep(c(0, 33.6, 67.2), each = 4)
yield <- 8.5 + 0.05 * nitro + rnorm(length(nitro), sd = 1.1)

nitrogen <- data.frame(nitro, yield)
nitrogen
```

First, we always inspect the data.

```{r, echo=TRUE}
ggplot(nitrogen, aes(x = nitro, y = yield)) +
  geom_point(pch = 21, size = 2.5) +
  labs(x = "Nitrogen rate",
       y = "Yield") +
  theme_linedraw()
```

The center of the points appears to increase approximately linearly with
nitrogen, so a simple linear model is reasonable as a first approach.

## Connecting to previous linear models

This part is important: you have seen this model structure before. Since
module 6, we have been using the same linear additive framework:

$${Y_{ij} = \mu + \tau_i + \varepsilon_{ij}}$$

where $\tau_i$ is a categorical treatment effect, a fixed shift in the
mean for each group. When we called `lm(yield ~ trt)`, R was fitting
exactly this model.

Simple linear regression uses that same framework, with one key change:
the predictor is **continuous** instead of **categorical**.

$${Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i}$$

Now $\beta_1 x_i$ replaces $\tau_i$: instead of a separate mean shift
for each group, the mean shifts continuously and proportionally as $x$
increases. The assumptions, the residual structure, and the hypothesis
test are structurally identical.

The parallel with previous modules is:

-   previous model: $Y_{ij} = \mu + \tau_i + \varepsilon_{ij}$
-   current model: $Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$
-   previous R call: `lm(yield ~ trt)` and `trt` is a **categorical**
    variable
-   current R call: `lm(yield ~ nitro)` and `nitro` is a **continuous**
    variable
-   previous null hypothesis: all $\tau_i = 0$
-   current null hypothesis: $\beta_1 = 0$

So this is not a new statistical world. It is an extension of the one
you already know.

## Linear equation

The regression equation is:

$${\hat{y} = \hat{\beta}_0 + \hat{\beta}_1x}$$

Where:

-   $\hat{\beta}_0$ is the intercept
-   $\hat{\beta}_1$ is the slope

Interpretation:

-   intercept: expected $y$ when $x = 0$
-   slope: expected change in $y$ for each one-unit increase in $x$

The diagram below annotates the fitted line directly to make both terms
concrete on the plot.

```{r}
m_ann   <- lm(yield ~ nitro, data = nitrogen)
a0_ann  <- coef(m_ann)[1]
a1_ann  <- coef(m_ann)[2]
y_at_34 <- a0_ann + a1_ann * 33.6

ggplot(nitrogen, aes(x = nitro, y = yield)) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue",
              linewidth = 0.8, fullrange = TRUE) +
  geom_point(pch = 21, size = 3) +
  annotate("point", x = 0, y = a0_ann, shape = 18, size = 5,
           color = "firebrick") +
  annotate("text", x = 3, y = a0_ann + 0.6,
           label = paste0("intercept = ", round(a0_ann, 2)),
           hjust = 0, color = "firebrick", size = 3.5) +
  annotate("segment", x = 0, xend = 33.6,
           y = a0_ann, yend = a0_ann,
           color = "darkgreen", linewidth = 0.7, linetype = "dashed") +
  annotate("segment", x = 33.6, xend = 33.6,
           y = a0_ann, yend = y_at_34,
           color = "darkgreen", linewidth = 0.9) +
  annotate("text", x = 35, y = (a0_ann + y_at_34) / 2,
           label = paste0("slope = ", round(a1_ann, 3), " per unit N"),
           hjust = 0, color = "darkgreen", size = 3.5) +
  xlim(-5, 80) +
  labs(x = "Nitrogen rate", y = "Yield") +
  theme_linedraw()
```

## Model assumptions

For the usual inference from `summary()` to be reliable, we still need
the same core assumptions:

-   $x$ is controlled by the researcher or measured with negligible
    error
-   the random errors are independent and have a common variance
-   the random errors are approximately normally distributed
-   the straight-line model is the correct functional form

These are the same assumptions you already checked with residual plots
in previous modules. The only difference is what we are modeling:
instead of asking whether group means differ, we are asking whether a
straight line captures how mean $y$ changes as $x$ changes.

One practical note on the intercept: interpreting $\hat{\beta}_0$ as the
expected response when $x = 0$ only makes sense when $x =
0$ is within or close to the observed range. If no observations were
taken near zero, the intercept is a mathematical anchor that positions
the line, not a directly interpretable quantity. In the nitrogen example
this is not a problem, because zero nitrogen is a real experimental
condition and within the observed range.

## Fitting the model in R

```{r, echo=TRUE}
model1 <- lm(yield ~ nitro, data = nitrogen)
summary(model1)
```

Worked interpretation example:

Suppose your model output shows an intercept near 8.5 and a slope near
0.05. A clear way to report that is: with zero nitrogen, expected yield
is about 8.5 units, and each additional 1 unit of nitrogen is associated
with an average increase of about 0.05 yield units. This does not mean
every single observation increases by exactly 0.05. It means the
*average trend* in the fitted line changes by that amount, while
individual observations still vary around that trend due to residual
variation.

Now we can overlay the fitted line.

```{r}
ggplot(nitrogen, aes(x = nitro, y = yield)) +
  geom_point(pch = 21, size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(x = "Nitrogen rate",
       y = "Yield") +
  theme_linedraw()
```

## Interpreting coefficients

From the fitted model:

-   the estimated intercept is the expected yield at zero nitrogen
-   the estimated slope is how much yield changes per unit nitrogen

We can also make predictions for values inside the observed range
(interpolation).

```{r, echo=TRUE}
coef(model1)

predict(model1, newdata = data.frame(nitro = c(15, 50)))
```

## Least-squares idea

The fitted line is chosen by least squares: it minimizes the sum of
squared residuals.

Residual for each point:

$${residual_i = y_i - \hat{y}_i}$$

Least squares chooses coefficients that minimize:

$${\sum (y_i - \hat{y}_i)^2}$$

The red segments in the plot below show each individual residual. The
fitted line is the one position that makes the sum of squared segment
lengths as small as possible.

```{r}
nitrogen_resid         <- nitrogen
nitrogen_resid$fitted  <- fitted(model1)

ggplot(nitrogen_resid, aes(x = nitro, y = yield)) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue",
              linewidth = 0.8) +
  geom_segment(aes(xend = nitro, yend = fitted),
               color = "firebrick", linewidth = 0.7) +
  geom_point(pch = 21, size = 3, fill = "white") +
  labs(x = "Nitrogen rate", y = "Yield",
       title = "Residuals as vertical segments") +
  theme_linedraw()
```

### A quick visualization of Least-squares

This is a useful place to visualize how least squares chooses a slope.

**Important note:** this is not the algorithm R uses internally for
`lm()`. R solves the normal equations analytically for this model class.
We are doing this only as a conceptual visualization. In the full
least-squares problem, both the intercept and slope are estimated
simultaneously. Here, we fix the intercept only to simplify
visualization and isolate how the slope affects the error.

From our fitted model, the intercept is about 8.48 and the slope is
about 0.05. We will keep the intercept fixed at 8.48 and try several
candidate slope values. For each slope, we compute the sum of squared
errors (SSE).

Now we plot the SSE for each candidate slope value. This gives the
characteristic U-shape, with the minimum at the least-squares estimate.

```{r}
b0_fix <- coef(model1)[1]
b1_hat <- coef(model1)[2]

b1_vals <- seq(0.01, 0.10, 0.001)
sse <- sapply(b1_vals, function(b1) {
  y_hat <- b0_fix + b1 * nitrogen$nitro
  sum((nitrogen$yield - y_hat)^2)
})

b1_candidates <- c(0.02, 0.04, 0.06, 0.08)
candidate_cols <- c("firebrick", "darkorange", "darkgreen", "steelblue")

par(mfrow = c(1, 2))
plot(nitrogen$nitro, nitrogen$yield,
     pch = 16, bg = "white",
     xlab = "Nitrogen rate", ylab = "Yield",
     main = "Four candidate slopes\n(intercept fixed)")

for (i in seq_along(b1_candidates)) {
  abline(a = b0_fix, b = b1_candidates[i],
         col = candidate_cols[i], lwd = 2)
}

abline(a = b0_fix, b = b1_hat, col = "black", lwd = 2, lty = 2)

legend("topleft",
       legend = c(paste0("b1 = ", b1_candidates),
                  paste0("best b1 = ", round(b1_hat, 4))),
       col = c(candidate_cols, "black"),
       lwd = 2, lty = c(rep(1, length(b1_candidates)), 2),
       bty = "n")

plot(b1_vals, sse, type = "l", lwd = 2,
     xlab = expression(beta[1]),
     ylab = "Sum of Squared Errors (SSE)",
     main = "SSE as a function of slope",
     ylim = c(5, 70))

points(b1_candidates,
       sapply(b1_candidates, function(b1) {
         sum((nitrogen$yield - (b0_fix + b1 * nitrogen$nitro))^2)
       }),
       pch = 19, col = candidate_cols)

best_idx <- which.min(sse)
points(b1_vals[best_idx], sse[best_idx], pch = 19, col = "black", cex = 1.2)

abline(v = b1_vals[best_idx], col = "black", lty = 2)
text(b1_vals[best_idx], 7,
     labels = paste0(" min at b1 = ", round(b1_vals[best_idx], 4)),
     pos = 1, cex = 0.9)

par(mfrow = c(1, 1))
```

## Significance of coefficients

In `anova(model1)`, the p-value for the slope tests:

-   $H_0: \beta_1 = 0$
-   $H_a: \beta_1 \ne 0$

If the p-value is small (for example, $< 0.05$), we have evidence that
the slope is not zero, so there is a meaningful linear trend between $x$
and $y$.

## Analysis of variance for regression

You can also view the same model test through ANOVA.

```{r, echo=TRUE}
anova(model1)
```

For simple linear regression, this ANOVA test is equivalent to testing
whether the slope differs from zero. So the logic is exactly the same as
previous modules. The predictor changed from categorical to continuous,
but the variation partition, F-statistic, and p-value interpretation
follow the same workflow you already know.

## Measuring fit with $R^2$

The model summary also reports $R^2$.

-   $R^2$ is the proportion of variation in $y$ explained by the fitted
    line
-   values closer to 1 indicate better fit
-   values near 0 indicate little explanatory value

A useful decomposition is:

$${SST = SSR + SSE}$$

where:

-   $SST$ = total variation around the mean
-   $SSR$ = variation explained by the model
-   $SSE$ = residual (unexplained) variation

So:

$${R^2 = \frac{SSR}{SST}}$$

In simple linear regression, $R^2$ is also equal to the square of the
Pearson correlation between observed and fitted values.

The short check below verifies this directly in R.

```{r, echo=TRUE}
y_obs <- nitrogen$yield
y_hat <- fitted(model1)

SST <- sum((y_obs - mean(y_obs))^2)
SSR <- sum((y_hat - mean(y_obs))^2)

R2_manual <- SSR / SST
R2_corr <- cor(y_obs, y_hat)^2

c(R2_manual = R2_manual,
  R2_corr = R2_corr,
  R2_summary = summary(model1)$r.squared)
```

Two cautions are worth keeping in mind:

-   a high $R^2$ does not prove the model is correct
-   a modest $R^2$ can still be scientifically useful in noisy
    biological systems if effect estimates are interpretable and
    diagnostics are acceptable

## Checking model assumptions

Before trusting inferences and predictions, we inspect residuals.

```{r, echo=FALSE}
par(mfrow = c(1, 2))
plot(model1, which = 1)
plot(model1, which = 2)
par(mfrow = c(1, 1))
```

What we want to see:

-   no strong curvature in residuals vs fitted
-   no obvious funnel shape, which would indicate unequal variance
    across levels of $x$
-   residuals approximately following a straight line in normal Q-Q plot

In this example, the data are clean and the diagnostics look reasonably
well-behaved. In real experiments, however, the residual vs. fitted plot
may reveal curvature (suggesting the relationship is not truly linear)
or a funnel shape (suggesting unequal variance across $x$). In those
cases, the transformation and GLM tools from Module 9 apply here exactly
as they did for ANOVA. The diagnostic workflow is the same.

### Reality check: fitting a line to curved data

To emphasize why diagnostics are not optional, imagine the same
consultant had tested nine nitrogen rates instead of three. Perhaps a
research institute encouraged her to run a more detailed trial to better
characterize the response curve. Here is what a more thorough trial
might show:

```{r, echo=FALSE}
set.seed(20260404)
nitro_full <- rep(seq(0, 160, length.out = 9), each = 4)

yield_full <- nlraa::SSquadp3xs(nitro_full, 8.5, 0.05, 120) + rnorm(length(nitro_full), sd = 0.2)

nitrogen_full <- data.frame(nitro = nitro_full, yield = yield_full)

ggplot(nitrogen_full, aes(x = nitro, y = yield)) +
  geom_point(pch = 21, size = 2.5) +
  labs(x = "Nitrogen rate",
       y = "Yield") +
  theme_linedraw()

```

```{r, echo=TRUE}
model_full <- lm(yield ~ nitro, data = nitrogen_full)
summary(model_full)
```

The summary looks impressive: a significant slope (p-value near 0) and
$R^2 \approx 0.82$.

Let's take a look at the fitted line:

```{r}
ggplot() +
  geom_point(data = nitrogen_full,
             aes(x = nitro, y = yield),
             pch = 21, size = 2.5) +
  geom_line(aes(x = nitrogen_full$nitro, y = fitted(model_full)),
            col = 'steelblue')+
  labs(x = "Nitrogen rate",
       y = "Yield") +
  theme_linedraw()
```

But look at the residuals:

```{r}
par(mfrow = c(1, 2))
plot(model_full, which = 1)
plot(model_full, which = 2)
par(mfrow = c(1, 1))
```

The residuals show a strong curved pattern. The line systematically
underpredicts at low and high nitrogen, but overpredicts in the middle.
A significant p-value and high $R^2$ do not guarantee the model fits
correctly. Only by inspecting residuals do we discover the true
relationship is nonlinear (likely quadratic), and a straight line is
inappropriate no matter how flat the summary statistics look.

## Anscombe's quartet: why plots matter

A classic reminder of why residual inspection matters comes from
[Anscombe (1973)](https://doi.org/10.1080/00031305.1973.10478966). He
constructed four datasets that are, for all practical purposes,
numerically identical:

-   same means and variances of $x$ and $y$
-   same fitted line: $\hat{y} = 3 + 0.5x$
-   same $R^2 = 0.667$
-   same F-statistic and p-value

But when plotted, only one dataset shows a clean linear relationship.
The others reveal curvature, a single influential outlier, and a
perfectly collinear set interrupted by one extreme point.

This dataset is built into base R.

```{r}
par(mfrow = c(2, 2))
for (i in 1:4) {
  xi <- anscombe[[paste0("x", i)]]
  yi <- anscombe[[paste0("y", i)]]
  plot(xi, yi,
       main  = paste("Dataset", i),
       xlab  = "x", ylab = "y",
       pch   = 19, xlim = c(4, 19), ylim = c(3, 13))
  abline(lm(yi ~ xi), col = "steelblue", lwd = 1.5)
}
par(mfrow = c(1, 1))
```

The takeaway is direct: identical numerical summaries can hide very
different underlying structures. A high $R^2$ and a significant p-value
do not substitute for looking at the data. This is exactly why we always
plot before and after fitting, and why the diagnostics in the previous
section are not optional.

# Extrapolation

Interpolation is prediction **within** the observed range of $x$.

Extrapolation is prediction **outside** that range.

Extrapolation can be risky because the true relationship may change
outside the sampled region.

In this example, we observed nitrogen rates from 0 to 67.2. Predicting
at 40 is interpolation. Predicting at 130 is extrapolation.

The shaded region below marks the observed range. The green point is
within it; the red point is well outside.

```{r}
a0_ext <- coef(model1)[1]
a1_ext <- coef(model1)[2]

ggplot(nitrogen, aes(x = nitro, y = yield)) +
  annotate("rect", xmin = 0, xmax = 67.2,
           ymin = -Inf, ymax = Inf, fill = "steelblue", alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue",
              linewidth = 0.8, fullrange = TRUE) +
  geom_point(pch = 21, size = 3) +
  annotate("point", x = 40, y = a0_ext + a1_ext * 40,
           shape = 17, size = 4, color = "darkgreen") +
  annotate("text", x = 40, y = a0_ext + a1_ext * 40 + 1.5,
           label = "x = 40\n(interpolation)",
           hjust = 0.5, color = "darkgreen", size = 3.2) +
  annotate("point", x = 130, y = a0_ext + a1_ext * 130,
           shape = 17, size = 4, color = "firebrick") +
  annotate("text", x = 130, y = a0_ext + a1_ext * 130 - 1,
           label = "x = 130\n(extrapolation)",
           hjust = 0.5, color = "firebrick", size = 3.2) +
  annotate("text", x = 33, y = 7.6,
           label = "observed range", color = "steelblue", size = 3.2) +
  xlim(-10, 175) +
  labs(x = "Nitrogen rate", y = "Yield") +
  theme_linedraw()
```

# Common mistakes to avoid

-   Treating correlation as proof of causation.
-   Reporting a significant p-value without interpreting effect size.
-   Interpreting a high $R^2$ as proof the model is correct.
-   Ignoring residual patterns because the slope is significant.
-   Concluding the relationship **is** linear just because the slope is
    significant: a curved trend can produce a non-zero linear component
    and a significant p-value even when the straight-line model is
    wrong.
-   Concluding there is **no** relationship just because the slope is
    non-significant: a curved relationship with zero linear slope will
    yield a non-significant p-value even though a real relationship
    exists. Always look at the scatter plot.
-   Extrapolating far outside sampled values as if uncertainty stayed
    the same.

In practice, good interpretation combines all pieces of evidence: effect
size, uncertainty, residual diagnostics, and biological plausibility.

# Summary

In this module, we learned:

-   Correlation quantifies association, not causation.
-   Regression models directional relationships and supports prediction.
-   The slope is the expected change in $y$ per unit change in $x$.
-   Least squares fits the line by minimizing squared residuals.
-   Significance of the slope can be tested with coefficient tests or
    ANOVA.
-   $R^2$ summarizes model fit.
-   Diagnostics are essential to check whether a linear model is
    appropriate.
-   Extrapolation should be approached cautiously.

The key conceptual shift in this module is moving from comparing groups
to modeling relationships. This allows us not only to detect
differences, but to quantify how variables change together in
biologically meaningful ways, and to make calibrated predictions for
values we did not directly observe.
